% This file defines the blessings to be inserted into haggadah.tex  In this way
% multiple formats can be typeset very quickly.  haggadah.tex should not normally
% need changing.  Note that this is not the most readable way to insert text
% into a LaTeX document, but it is the most powerful: the macros defined here
% are directly excecuted when building the document.

%!TEX TS-program = lualatex 
%!TEX encoding = UTF-8 Unicode
%\usepackage{fontspec}
%\usepackage{polyglossia}
%\setdefaultlanguage{english}
%\setotherlanguages{czech, hebrew}
%thia line is a bug \renewfontfamily{\hebrewfont}{Ezra SIL SR Regular}
% 
\usepackage{paracol}

\newcommand{\CandlesBlessingHeb}{
	\begin{center}
			%\textdir TRT
			\texthebrew{\LARGE בָּרוּךְ אַתָּה אֲ-דֹנָי אֱ-לֹהֵינוּ מֶלֶךְ הָעוֹלָם, אֲשֶׁר קִדְּשָׁנוּ בְּמִצְוֹתָיו	}\\\texthebrew{\Huge וְצִוָּנוּ לְהַדְלִיק נֵר שֶׁל יוֹם טוֹב}
	\end{center}
\vspace*{2ex}
}
%
\newcommand{\CandlesBlessingTlit}{%
	Baruch atah Adonai eloheinu melech ha-olam asher kidd’shanu
	b’mitzvotav v’tzivvanu l’hadlik ner shel %(shabbat v’shel) 
	yom tov.}
%
\newcommand{\CandlesBlessingEng}
	{ May you be blessed, Lord our G-d, ruler of the universe, who sanctifies us via his commandments and who commands us to light the lights of %(Shabbat and)
		 festivals.}
%
\newcommand{\CandlesBlessingCz}{
Požehnaný jsi Ty, Hospodine, Bože náš, Králi všehomíra, jenž nás posvětil Svými přikázáními a přikázal nám zapálit světlo %(šabatové a) 
sváteční.}
%
%
\newcommand{\WineBlessingHeb}{
	\begin{center}
	%\textdir TRT
	\texthebrew{\LARGE בָּרוּךְ אַתָּה אֲ-דֹנָי אֱ-לֹהֵינוּ מֶלֶךְ הָעוֹלָם}\\
	\texthebrew{\Huge בּוֹרֵא פְּרִי הַגָּפֶן}
	\end{center}
	\vspace*{2ex}
	}
%
\newcommand{\WineBlessingTlit}{%
		Baruch atah Adonai eloheinu melech ha-olam borei p’ri ha-gafen.}
%		
\newcommand{\WineBlessingEng}
		{We thank You Lord our God, ruler of the universe, for making the fruit of the vine.}
%		
\newcommand{\WineBlessingCz}{Požehnaný jsi Ty, Hospodine, Bože náš, Králi všehomíra, stvořiteli plodů vinné révy.}
%
%
\newcommand{\ShehechyanuBlessingHeb}{	
	\begin{center}
		%\textdir TRT
		\texthebrew{\LARGE בָּרוּךְ אַתָּה אֲ-דֹנָי אֱ-לֹהֵינוּ מֶלֶךְ הָעוֹלָם}\\
		\texthebrew{\Huge שֶׁהֶחֱיָנוּ וְקִיְּמָנוּ וְהִגִּיעָנוּ לִזְמַן הַזֶּה}
	\end{center}
	\vspace*{2ex}
	}
\newcommand{\ShehechyanuBlessingEng}{%
	We praise You, Eternal God, Ruler of the universe, for keeping us alive,
	sustaining us, and enabling us to reach this season.
}
%
\newcommand{\ShehechyanuBlessingTlit}{%
	Baruch atah Adonai eloheinu melech ha-olam Shehechyanu V'kiymanu v'higiyanu lazman ha-zeh.
}
% 
\newcommand{\ShehechyanuBlessingCz}{Požehnaný jsi Ty, Hospodine, Bože náš, Králi všehomíra, že jsi nám dopřál dožít se, dočkat se a dosáhnout tohoto času.}
%
%
\newcommand{\KarpasBlessingHeb}{	
	\begin{center}
		%\textdir TRT
		\texthebrew{\LARGE בָּרוּךְ אַתָּה אֲ-דֹנָי אֱ-לֹהֵינוּ מֶלֶךְ הָעוֹלָם}\\
		\texthebrew{\Huge בּוֹרֵא פְּרִי הָאֲדָמָה}
	\end{center}
	\vspace*{2ex}
}
%
\newcommand{\KarpasBlessingTlit}{%
	Baruch atah Adonai eloheinu melech ha-olam borei p’ri ha-adamah.
}
%
\newcommand{\KarpasBlessingEng}{%
	We thank You God, for making food grow from the ground.
}
% 
\newcommand{\KarpasBlessingCz}{Požehnaný jsi Ty, Hospodine, Bože náš, Králi všehomíra, stvořiteli plodů země.}
%
\newcommand{\NetilatYadayimHeb}{
	\begin{center}
		%\textdir TRT
		\texthebrew{\LARGE בָּרוּךְ אַתָּה אֲ-דֹנָי אֱ-לֹהֵינוּ מֶלֶךְ הָעוֹלָםאֲ שֶּ ר קִ ד ש ָּ נוְּ ּבמִ צֹו תָּ יו,}\\
		\texthebrew{\Huge  וְצִוָּנוּ עַל נְטִילַת יָדַיִם}
	\end{center}}
\newcommand{\NetilatYadayimTlit}{
	Baruch Atah Adonai, Eloheinu Melech ha-olam, asher kid’shanu b’mitzvotav
	v’tzivanu al n’tilat yadayim.}
\newcommand{\NetilatYadayimEng}{Blessed are You, Lord our G-d, Ruler of the Universe, who sanctifies us with his commandments and commands us regarding the washing of hands.}
\newcommand{\NetilatYadayimCz}{Požehnaný jsi Ty, Hospodine, Bože náš, Králi všehomíra, % !!!!!!jenž nás posvětil Svými přikázáními a přikázal nám přikázání ohledně umývání rukou. OR:
že jsi nás posvětil svými příkazy a dal jsi nám příkaz o omývání rukou. }
%	
\newcommand{\HaMotziBlessingHeb}{	
	\begin{center}
		%\textdir TRT
		\texthebrew{\LARGE בָּרוּךְ אַתָּה אֲ-דֹנָי אֱ-לֹהֵינוּ מֶלֶךְ הָעוֹלָם}\\
		\texthebrew{\Huge הַמּוֹצִיא לֶחֶם מִן הָאָרֶץ}
	\end{center}
	\vspace*{2ex}
}
\newcommand{\HaMotziBlessingTlit}{%
	Baruch ata Adonai Eloheinu melech ha’olam hamotzi lechem min ha’aretz.}
%
\newcommand{\HaMotziBlessingEng}{%
	Blessed are you, Lord our God, ruler of the universe who brings forth bread from the earth.}
%
\newcommand{\HaMotziBlessingCz}{Požehnaný jsi Ty, Hospodine, Bože náš, Králi všehomíra, % původce chleba ze země/kdo nese....který neseš... chléb ze země/přináší chléb ze země? že vyvádíš ze země chléb.
jenž dáváš chlebu vyrůst ze země.}

	
\newcommand{\MatzahBlessingHeb}{
	\begin{center}
	\texthebrew{\LARGE בָּרוּךְ אַתָּה אֲ-דֹנָי אֱ-לֹהֵינוּ מֶלֶךְ הָעוֹלָםאֲ שֶּ ר קִ ד ש ָּ נוְּ ּבמִ צֹו תָּ יו,}\\
	\texthebrew{\Huge וְצִוָּנוּ עַל אֲכִילַת מַצָּה}
	\end{center}}
\newcommand{\MatzahBlessingTlit}{%
	Baruch Atah Adonai, Eloheinu Melech ha-olam, asher kid’shanu b’mitzvotav
	v’tzivanu al achilat matzah.}
\newcommand{\MatzahBlessingEng}{%
	We praise God, Ruler of Everything, who makes us holy through his commandments, and commands us to eat matzah.}
\newcommand{\MatzahBlessingCz}{Požehnán jsi, Hospodine, Bože náš, Králi světa, % že jsi nás posvětil svými přikázáními a přikázal nám přikázání jíst maca. OR Požehnán jsi, Hospodine, Bože náš, králi světa, 
	že jsi nás posvětil svými příkazy a dal jsi nám příkaz o pojídání nekvašených chlebů. }
%
%
\newcommand{\MarorBlessingHeb}{
	\begin{center}
			\texthebrew{\LARGE בָּרוּךְ אַתָּה אֲדֹנָי אֱלֹהֵינוּ מֶלֶךְ הָעוֹלָם אֲשֶׁר קִדְּשָׁנוּ בְּמִצְוֹתָיו,}\\
			\texthebrew{\Huge וְצִוָּנוּ עַל אֲכִילַת מָרוֹר}
	\end{center}}
%
\newcommand{\MarorBlessingTlit}{%
	Baruch Atah Adonai, Eloheinu Melech ha-olam, asher kid’shanu b’mitzvotav
v’tzivanu al achilat maror.}
%
\newcommand{\MarorBlessingEng}{%
	We praise God, Ruler of Everything, who made us holy through obligations, commanding us to eat bitter herbs.}
%
\newcommand{\MarorBlessingCz}{jenž nás posvětil Svými přikázáními a přikázal nám přikázání jíst hořké byliny.}
%
%
\newcommand{\BareichEng}{
\raggedright
\textit{We will say the Birkat Hamazon (Grace after meals) with some extra parts:}
\par\vspace*{2ex}
	\columnratio{0.3}
	\begin{paracol}{2}
		Birkat Hazan: \vspace*{4mm} \switchcolumn \vspace*{-1mm} We praise God for sustaining life and providing food for all creatures.\\ %(sung)
		%{\textcolor{darkblue} We praise God, Ruler of Everything, whose goodness sustains the world. You are the origin of love and compassion, the source of bread for all. Thanks to You, we need never lack for food; You provide food enough for everyone. We praise God, source of food for everyone.}
		\hspace*{50mm}{\color{black} Amen}
		%
		\switchcolumn \vspace*{2ex} Birkat HaAretz: \vspace*{6mm}	\switchcolumn  \vspace*{-1mm} We thank God for being compassionate and nourishing us with food and Torah.\\ %recount Exodus to Canaan. 
		%{\textcolor{darkblue} As it says in the Torah: When you have eaten and are satisfied, give praise to your God who has given you this good earth. We praise God for the earth and for its sustenance, and for the words of Torah which are sweet like honey on the tongue and a lamp to light our feet (ps: 119)} \textcolor{black}\\
		\hspace*{50mm}{\color{black} Amen}
		%{\textcolor{darkblue} Just as God sustained the Jewish people in the desert, so too God sustains us now and in the future. }\\
		%\textcolor{black}
		%\hspace*{80mm}{\large Amen}\vspace*{2ex}\\
		%
		\switchcolumn \vspace*{0.5ex} Birkat Yerushalayim: \vspace*{4mm} \switchcolumn  \vspace*{-1mm}	We ask God for His mercy and continued support.\\ %, and traditionally to quickly rebuild Jerusalem.
		\hspace*{50mm}{\color{black} Amen}\vspace*{2ex}
		%
		\switchcolumn \vspace*{1ex} Birkat HaTov VeHaMeitiv: \vspace*{4mm}	\switchcolumn  \vspace*{-1mm}	We remember the blessings God bestows upon us. \\%and request God never denies us anything good.\\
		\hspace*{50mm}{\color{black} Amen}\vspace*{2ex}
		%
		\switchcolumn Harahaman : \vspace*{4mm}	\switchcolumn  \vspace*{-1mm}	%\textit{(May the compassionate one)}\\
		We ask God to uplift us and our community, and all the world.\\
		%{\textcolor{darkblue} Renew our spiritual center in our time. We praise God, who centers us. May the source of peace grant peace to us, to the Jewish people, and to the entire world.}\textcolor{black}\\
		\hspace*{50mm}{\color{black} Amen}
		\vspace*{2ex}
	\end{paracol}
	{\textcolor{darkblue} \centering Spanish and Portugese Jews sing a Ladino song entitled “Bendigamos” or “We will bless” (let's try to find it on spotify).}
	}
%
\newcommand{\MiriamBlessing}{
	\begin{center}
		\large Zot Kos Miryam, kos mayim hayim. Zeikher l’yitziat Mitztrayim.
		\end{center}	
	This is the Cup of Miriam, the cup of living waters. 
	\vspace*{2ex}
	These are the living waters, God’s gift to Miriam, which gave new life to Israel as we struggled with ourselves in the wilderness. Blessed are You God, Who brings us from the narrows into the wilderness, sustains us with endless possibilities, and enables us to reach a new place.”
	}
\newcommand{\NirtzahBlessing}{\centering \color{darkblue} \largehebrew{לְשָׁנָה הַבָּאָה בִּירוּשָלָיִם}\\
	L’shana haba-ah biy’rushalayim\\
	\color{black} příští rok v Jeruzalémě\\
	\color{midblue} NEXT YEAR IN JERUSALEM!\\}

