% This file defines the signposting text to be inserted into haggadah.tex  In this way
% multiple formats can be typeset very quickly.  haggadah.tex should not normally
% need changing.  Note that this is not the most readable way to insert text
% into a LaTeX document, but it is the most powerful: the macros defined here
% are directly excecuted when building the document.

%\usepackage{fontspec}
%\usepackage{polyglossia}
%\setdefaultlanguage{english}
%\setotherlanguages{czech, hebrew}
%\newfontfamily{\hebrewfont}{New Peninim MT}

% For the title page
\newcommand{\contents}{%

\newcommand{\contentsTranslation}{%
}
% 
\newcommand{\Neirot}{
{We normally begin after Havdalah by lighting the festival candles, but today we will come back to this once the sun has set fully, so we can sleep on time}
\CandlelightingBlessing 
}

\newcommand{\Kaddeish}{%
	\newcommand{\wineblessing}{%
		{%
			\texthebrew{שלום עולם
ברָּ וך ְּ אַ תָּ ּה יי ,ָּ אֱֹלהֵ ינו מֶּ לךֶּ ְּ הָּ עֹולם,ְָּּ בוֹּ רֵ א פ רִ י הַ גפֶָּּ ן}
}
{%
Baruch atah Adonai eloheinu melech ha-olam borei p’ri ha-gafen.
We thank You God, for making the fruit of the vine
  }
  \per
}
\per}

\newcommand{\Shehechyanu}

We praise You, Eternal God, Ruler of the universe, for keeping us alive,
sustaining us, and enabling us to reach this season.

\newcommand{\urchatz}{%
  \l{%
latin
  }
  \e{%
english
  }
}

% 
\newcommand{\Karpas}{%
	dip parsley in salt water to remember tears and spring
Baruch atah Adonai eloheinu melech ha-olam borei p’ri ha-adamah.
We thank You God, for making food grow from the ground.
}
\newcommand{\KarpasTranslation}{%
the translated blessing
}
% 
\newcommand{\Yachatz}{%
  \l{%

middle matzah broken, ()larger piece LATER hidden, smaller replaced) all lifted while the following aramaic is read:

Ha lachma anya di achalu
av’hatana b’ar’ah d’mitzrayim.
Kol dich’fin yeiy’tei v’yeichul
Kol ditz’rich yeiy’tei v’yif’sach.
Ha-shata hacha –
l’shata d’atya b’ar’ah d’yisra’el
Ha-shata avdei –
l’shata d’atya b’nei chorin
  }
  \e{%
This is the bread of affliction our ancestors ate in the land of Egypt. Let all who
are hungry come and eat; let all who are in need come and share our Passover.
This year here, next year in the land of Israel; this year oppressed, next year free.
  }
}
\newcommand{\MaNishtana}{%
Mah nishtanah ha-lailah ha-zeh mi-kol ha-leilot!
She-b’chol ha-leilot anu ochlin chameitz u-matzah, ha-lailah ha-zeh kulo matzah?
She-b’chol ha-leilot anu ochlin sh’ar y’rakot, ha-lailah ha-zeh maror?
She-b’chol ha-leilot ein anu matbilin afilu pa’am achat, ha-lailah ha-zeh sh’teif’amim?
She-b’chol ha-leilot anu ochlin bein yoshvin u-vein m’subin, ha-lailah ha-zeh
kullanu m’subin?

How different this night is from all other nights!
On all other nights we eat either leavened or unleavened bread…why only unleavened bread tonight?
On all other nights we eat different types of herbs and vegetables…why bitter herbs tonight?
On all other nights we do not even dip once…why do we dip twice tonight?
On all other nights we eat either sitting or leaning…why do we all lean tonight?
}
\newcommand{\MaNishtanaTranslation}{%
The souls of the just are in the hand of God, and no torment shall touch them. They seemed, in the view of the foolish, to be dead: but they are in peace. Alleluia.
}
\newcommand{\Maggid}{%
  \l{%
passage or poem for story
  }
  \e{%
We present to thee, O Lord, these gifts for our sacrifice, as honoring thy holy ones may they please thee, and of thy pity, may they become profitable to us.
  }
}
\newcommand{\Dayyeinu}{%
image of plagues
We name the plagues and spill a drop of drink on each one, like a tear for the suffering they caused in the world.
	\begin{itemize}
	\color{sectioncolour}
	\item Dam
	\item Tz'farde'a
	\item Kinim
	\item A'rov
	\item Dever
	\item Sh'chin
	\item Barad
	\item Ar'beh
	\item Choshech
	\item Makkat b'chorot
\end{itemize}

Many are the things HaShem did for the sake of our ancestors. Any one of these things by itself would have sufficed: Dayyeinu.

	\begin{itemize}
		If G-d had brought us out of Egypt and not given us Shabbat : Dayyeinu
	\end{itemize}

}
\newcommand{\communionTranslation}{%
Blessed are the clean of heart, for they shall see God. Blessed are the peacemakers, for they shall be called children of God. Blessed are they who suffer persecution for justice’ sake, for theirs is the kingdom of heaven.
}
\newcommand{\Hallel}{%
  \l{%
    Hallelu hallelu hallelu, hallelu, halleluyah!
    Kol ha-n’shamah t’hallel yah, hallelu halleluyah!
    Let us give praise – Let us all praise God. Halleluyah!
  }
  \e{%
second glass: \wineblessing
  }
  \per
}



% File paths: we don't use symlinks as (a) not all platforms support them, and
% (b) they don't fit nicely with the flow we're using.

%\newcommand{\kyriePath}{../Ordinaries/masses/4/kyrie}
%\newcommand{\gloriaPath}{../Ordinaries/masses/4/gloria}
%\newcommand{\sanctusPath}{../Ordinaries/masses/4/sanctus}
%\newcommand{\agnusPath}{../Ordinaries/masses/4/agnus}
%\newcommand{\itePath}{../Ordinaries/masses/4/ite}
% 
%\newcommand{\creedPath}{../Ordinaries/credo/1/credo}
% 
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "missalette"
%%% End:
